Extracting Angular Observables without a Likelihood\\and Applications to Rare Decays

### 2015-03-10 draft.tex
first bunch of changes

%% Draft Macros %%
\newcommand{\todo}[1]{{\color{red}\bf ToDo: #1}}
\newcommand{\danny}[1]{{\color{purple}#1}}
\newcommand{\fred}[1]{{\color{brown!85!black}#1}}
\newcommand{\citeneeded}{{\color{red}\bf [cite needed]}}
\newcommand{\add}[1]{\textcolor{green!85!black}{#1}}
\newcommand{\del}[1]{\textcolor{red!85!black}{#1}}

@@

Our initial motivation for studying what we wish to call the
\emph{method of moments} is the determination of angular observables
in the rare FCNC-mediated decay $\bar{B}\to \bar{K}^*(\to
\bar{K}\pi)\ell^+\ell^-$. However, the method we describe in the
following is general and applies to \add{all decay or scattering
processes that can be formulated in terms of an orthogonal basis of angular functions.}
\fred{Not really arbitrary, it only works with orthogonal
basis functions.} \danny{Is this OK?}
We find a previous work \cite{Dighe:1998vk} that
advocates this method chiefly for the determination of angular
observables in non-leptonic $B$ decays but also mentions the
applicability to semileptonic decays. Our aim is to improve upon this
previous work by studying the uncertainties that are introduced by
mismodeling of the angular distribution, and by working out a recipe
to determine and unfold detector effects. The latter is crucial for
the application of the method to real data. We show that
the method of moments has several major advantages over the usual approach
based on likelihood fits:
\begin{enumerate}
    \item Likelihood fits have convergence problems for a small number of
        events, and can require reparametrizations and/or approximations
        for a successful fit to the signal PDF. \add{As an example,
        see the LHCb analysis of \dots}.
        The method of moments does not require reparametrizations or approximations.

@@

\fred{For us, the major advantage is that the experiment can obtain all angular observables at once, and if there are more, one doesn't need more data. I'm not sure this becomes clear from the text, point 1.}

@@

\item The estimate is unbiased for any $N$. In contrast, the
  maximum-likelihood estimate has a bias of order
  $1/N$~\cite{Cox:1968}. In practice, one should keep in mind the
  bias-variance trade-off: it is a well known phenomenon that removing
  the bias usually leads to an increase in variance of the sampling
  distribution of the estimator and vice versa~\cite[sec. 7.3]{James:2006zz}. From a
  Bayesian decision-theory point of view, both contribute similarly to
  the expected loss associated with deciding on just one value of the
  unknown parameter. \add{One should therefore not prefer the method of moments
  over likelihood fits just because the former reduces}
  the bias~\cite[sections 13.8,17.2]{jaynes:2003}. In fact, for
  the results discussed below in \refsec{numerics}, the likelihood
  fits --- if they converge --- produce uncertainties \add{(10--15)\%}
      smaller than those from the method of moments  \fred{Is this true, Marcin?} \danny{Marcin already confirmed!}.
\item The approximate multivariate Gaussian distribution of $\vecest{S}$
    allows easier and more \add{precise} transfer of the information in the
  data to interested theorists for more accurate fits of
  standard-model and new-physics parameters \cite{Altmannshofer:2013foa,
Descotes-Genon:2013wba,Beaujean:2013soa},
  or for more precise predictions of optimized observables; see e.g.,
  \cite{Egede:2008uy,Egede:2010zc,Bobeth:2010wg,Becirevic:2011bp,
    Bobeth:2012vn,Matias:2012xw,DescotesGenon:2012zf} for definitions
  of such optimized observables in $B\to K^*\ell^+\ell^-$ decays,
  \cite{Faller:2013dwa} for application to the decay $B\to
  \pi\pi\ell^-\bar\nu_\ell$, and \cite{Boeer:2014xx} for observables
  in $\Lambda_b\to\Lambda(\to N\pi)\ell^+\ell^-$). While the
  likelihood also approaches a multivariate Gaussian as $N \to
  \infty$, the two methods differ in their utility as input for
  theorists if $\widehat{\vec{S}}$ is not well
  inside the physical region. For example, suppose there are two
  angular observables that are constrained to a triangular region by
  phase-space or unitarity arguments as
  \begin{equation}
    \label{eq:constr-ex}
    |S_1| + S_2 \le 1, \, S_1 \in [-1,1],\, S_2 \in [0,1] \,.
  \end{equation}
  It may (and often does) happen in practice that $\widehat{\vec{S}}$
  is close or even outside the allowed region such that a significant
  part of the probability mass covers unphysical values. In a Bayesian
  fit, one would take $\mathcal{N}(\vecest{S} | \vec{S},
  \widehat{\Sigma})$ as the sampling distribution of the ``data'', and
  simply set a uniform prior on the triangle in the $\vec{S}$ plane
  defined by \refeq{constr-ex} to have a well defined problem. This
  could be trivially combined with other independent information in a
  global fit. Someone with a different physics model might have to
  consider a
  different physical region, and could incorporate it just as easily.\\

  For a likelihood fit, the constraint needs to be part of the
  analysis performed by the experimental collaboration, and the
  resulting likelihood as a function of $\vec{S}$ is distinctly not
  Gaussian.  Communicating such a result has
  proved to be challenging due to technical reasons (such as data formats,
  size etc.) \add{as well as political reasons; i.e., the publication of
  the likelihood also publishes collaboration-internal data.}
  This leads to the
  undesirable situation that only the mode and standard errors
  are reported, and theorists often include the results as independent
  measurements with a Gaussian distribution and disregard the boundary
  problem as well as correlations altogether.

>>>>

%% Draft Macros %%
\newcommand{\todo}[1]{{\color{red}\bf ToDo: #1}}
\newcommand{\danny}[1]{{\color{purple}#1}}
\newcommand{\fred}[1]{{\color{brown!85!black}#1}}
\newcommand{\marcin}[1]{{\color{cyan}#1}}
\newcommand{\citeneeded}{{\color{red}\bf [cite needed]}}
\newcommand{\add}[1]{\textcolor{green!85!black}{#1}}
\newcommand{\del}[1]{\textcolor{red!85!black}{#1}}

@@

Our initial motivation for studying what we wish to call the
\emph{method of moments} is the determination of angular observables
in the rare FCNC-mediated decay $\bar{B}\to \bar{K}^*(\to
\bar{K}\pi)\ell^+\ell^-$. However, the method we describe in the
following is general and applies to all decay or scattering
processes that can be formulated in terms of an orthogonal basis of angular functions.
We find a previous work \cite{Dighe:1998vk} that
advocates this method chiefly for the determination of angular
observables in non-leptonic $B$ decays but also mentions the
applicability to semileptonic decays. Our aim is to improve upon this
previous work by studying the uncertainties that are introduced by
mismodeling of the angular distribution, and by working out a recipe
to determine and unfold detector effects. The latter is crucial for
the application of the method to real data. We show that
the method of moments has several major advantages over the usual approach
based on likelihood fits:
\begin{enumerate}
    \item Likelihood fits have convergence problems for a small number of
        events, and can require reparametrizations and/or approximations
        for a successful fit to the signal PDF. \add{As an example,
        see the LHCb analysis of \dots}.
        The method of moments does not require reparametrizations or approximations.

@@

\fred{For us, the major advantage is that the experiment can obtain all angular observables at once, and if there are more, one doesn't need more data. I'm not sure this becomes clear from the text, point 1.}\marcin{I think it is ok}

@@

\item The estimate is unbiased for any $N$. In contrast, the
  maximum-likelihood estimate has a bias of order
  $1/N$~\cite{Cox:1968}. In practice, one should keep in mind the
  bias-variance trade-off: it is a well known phenomenon that removing
  the bias usually leads to an increase in variance of the sampling
  distribution of the estimator and vice versa~\cite[sec. 7.3]{James:2006zz}. From a
  Bayesian decision-theory point of view, both contribute similarly to
  the expected loss associated with deciding on just one value of the
  unknown parameter. One should therefore not prefer the method of moments
  over likelihood fits just because the former reduces
  the bias~\cite[sections 13.8,17.2]{jaynes:2003} \marcin{we said we cahnge this but after reading it again I like it and would not change}. In fact, for
  the results discussed below in \refsec{numerics}, the likelihood
  fits --- if they converge --- produce uncertainties (10--15)\%
      smaller than those from the method of moments.
\item The approximate multivariate Gaussian distribution of $\vecest{S}$
    allows easier and more \add{precise} transfer of the information in the
  data to interested theorists for more accurate fits of
  standard-model and new-physics parameters \cite{Altmannshofer:2013foa,
Descotes-Genon:2013wba,Beaujean:2013soa},
  or for more precise predictions of optimized observables; see e.g.,
  \cite{Egede:2008uy,Egede:2010zc,Bobeth:2010wg,Becirevic:2011bp,
    Bobeth:2012vn,Matias:2012xw,DescotesGenon:2012zf} for definitions
  of such optimized observables in $B\to K^*\ell^+\ell^-$ decays,
  \cite{Faller:2013dwa} for application to the decay $B\to
  \pi\pi\ell^-\bar\nu_\ell$, and \cite{Boeer:2014xx} for observables
  in $\Lambda_b\to\Lambda(\to N\pi)\ell^+\ell^-$). While the
  likelihood also approaches a multivariate Gaussian as $N \to
  \infty$, the two methods differ in their utility as input for
  theorists if $\widehat{\vec{S}}$ is not well
  inside the physical region. For example, suppose there are two
  angular observables that are constrained to a triangular region by
  phase-space or unitarity arguments as
  \begin{equation}
    \label{eq:constr-ex}
    |S_1| + S_2 \le 1, \, S_1 \in [-1,1],\, S_2 \in [0,1] \,.
  \end{equation}
  It may (and often does) happen in practice that $\widehat{\vec{S}}$
  is close or even outside the allowed region such that a significant
  part of the probability mass covers unphysical values. In a Bayesian
  fit, one would take $\mathcal{N}(\vecest{S} | \vec{S},
  \widehat{\Sigma})$ as the sampling distribution of the ``data'', and
  simply set a uniform prior on the triangle in the $\vec{S}$ plane
  defined by \refeq{constr-ex} to have a well defined problem. This
  could be trivially combined with other independent information in a
  global fit. Someone with a different physics model might have to
  consider a
  different physical region, and could incorporate it just as easily.\\

  For a likelihood fit, the constraint needs to be part of the
  analysis performed by the experimental collaboration, and the
  resulting likelihood as a function of $\vec{S}$ is distinctly not
  Gaussian.  Communicating such a result has
  proved to be challenging due to technical reasons (such as data formats,
  size etc.) %\add{as well as political reasons; i.e., the publication of
 % the likelihood also publishes collaboration-internal data.}
  This leads to the
  undesirable situation that only the mode and standard errors
  are reported, and theorists often include the results as independent
  measurements with a Gaussian distribution and disregard the boundary
  problem as well as correlations altogether.
